---
title: "Using the GDC API"
author: "Sean Davis & Martin Morgan"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: tango
    df_print: paged
      
vignette: >
  %\VignetteIndexEntry{GenomicDataCommons API reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE,results='hide'}
library(knitr)
library(GenomicDataCommons)
library(DT)
opts_chunk$set(cache=TRUE)
```

# The API { .tabset .tabset-fade }

## Metadata Endpoints { .tabset .tabset-fade }

### General Concepts

The Projects, Files, Cases, and Annotations endpoints are are all
accessed using functions of the same name (all lowercase). The default
behavior of each is to query the GDC and return the first 10 records
with a default set of fields. In this section, the results are simply
converted to a tabular form, but the data are not always tabular and
may have one-to-many relationships, so this represents a
simplification. Accessing the data more directly will be covered in
later sections.

#### Querying

Parameter | Default | Description
--------- | :-----: | ---------------------------------------------------
filters	| null | Specifies search parameters
format | JSON  | Specifies the API response format: JSON, XML, or TSV
pretty | false | Returns response with indentations and line breaks in a human-readable format
fields | null | Specifies which fields to include in the response
size | 10 | Specifies the number of results to return
from | 1 | Specifies the first record to return from a set of search results
sort | null | Specifies sorting for the search results
facets | null | Provides all existing values for a given field and the number of records having this value.

#### Searching and filtering


Operator | Description | Number of Operands | Logic example
:---------:|----------------|:-------------:|-------------------------------
=  | equals (string or number) | one  | gender = "female"
!= | does not equal (string or number)  | one | project_id != "TARGET-AML"
<  | less than (number)  | one	| age at diagnosis < 90y
<=	| less than or equal (number)  | one | age at diagnosis <= 17
> | greater than (number)  | one | age at diagnosis > 50
>=	| greater than or equal (number) | one | age at diagnosis >= 18
is | is (missing) | one	| gender is missing
not	| not (missing)	| one | race not missing
in | matches a string or number in (a list)	| multiple | primary_site in [Brain, Lung]
exclude	| does not match any strings or values in (a list) | multiple | experimental_strategy exclude [WXS, WGS, "Genotyping array"]
and	| (operation1) and (operation2)	| multiple | {primary_site in [Brain, Lung]} and {gender = "female"}
or | (operation1) or (operation2) | multiple | {project_id != "TARGET-AML"} or {age at diagnosis < 90y}

#### The mapping endpoint

The mapping endpoint returns the available fields and the types of data contained in them for 
each of the four primary metadata endpoints. A data frame with is returned when the `mapping()` 
function is called with one of "cases", "files", "projects", or "annotations". In addition to the 
column definitions (first 5 columns), four additional columns are returned that allow subsetting
fields to those of interest. The endpoint-agnostic field names provided by the mapping endpoint
are compatible with the filters parameter but are not always compatible with the fields parameter.

mapping key | Description of value
----------- | --------------------------------------------------
defaults | The default set of fields included in the API response when the fields parameter is not used in the request 
expand | Field group names for use with the expand parameter 
fields	| All available fields in an endpoint-specific format that is compatible with both the filters and fields parameters 
multi | GDC internal use 
nested  | Nested fields

As an example, we can look at the fields associated with the [files][] endpoint.

```{r mapping10}
a = mapping('files')
colnames(a)
head(a)
# get the "default" fields
subset(a,defaults)
# get the "default" fields as character
subset(a,defaults)$field
```


#### Faceting (counting)

```{r facetProjectByProgram}
res = cases(facets='diagnoses.tumor_stage')
```

### Projects

To get a sense of what is available in the default Project entities,
we simply call the `projects()` function without any arguments. By
default, the first 10 projects are returned. The returned value for
projects is a `data.frame` since the Project entity is tabular. 

```{r project10}
a = projects()
datatable(a)
```

### Files

```{r files10}
library(jsonlite)
fields <- c("data_type", "updated_datetime", "file_name", "data_format", "access",
            "data_category", "file_size", "type")
a = files(fields=fields)
a = files(fields=fields,filters=make_filter('data_format'=='VCF'))
# Convert to dataframe
files_df = do.call(rbind, a)
datatable(files_df[,c('file_name','data_format','access','type','data_category')],rownames=FALSE)
```

### Cases

```{r cases10}
a = cases()
a             # 10 cases by default
lengths(a)    # cases have different elements
a[1]          # summary
head(a[[1]])  # values
str(a[[1]])
```

### Annotations

Annotations are notes added to individual cases, samples or
files. They can be searched and filtered in the same way as other
entities. See
[here](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Annotations/) for
details from Data Portal about annotations.

```{r annotations10}
a = annotations()
annots_df = do.call(rbind, a)
datatable(annots_df[,c('category','classification','notes','entity_type')],rownames=FALSE)
```

### Advanced querying with filters

The GDC
supports
[advanced queries](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Advanced_Search/) based
on filters and facets. The GenomicDataCommons package allows writing
queries using standard R syntax, simplifying the creation of
complicated filtering patterns. Examples follow:

```{r filtercases}
# youngcases = cases(filters=filters(diagnoses.age_at_diagnosis <= 10*365,endpoint='cases'))
```

## Data Access { .tabset .tabset-fade }

### Single or multiple small file downloads

```{r}
a = files(filters=make_filter('data_format'=='MAF' & 'access'=='open'))
fpaths = gdcdata(uuids=names(a),progress=FALSE)
basename(fpaths)
```

### Large or many files with GDC download tool

For many files or large files, the [GDC Download Tool][] can be more efficient and 
allows for download restarts if necessary.

```{r}
manifest_path = manifest(uuids = names(a))
manifest_path
```

Use the [GDC Download Tool][] for your
platform and to read and then download files in the `manifest_path` from above.  

## SessionInfo

```{r sessionInfo}
sessionInfo()
```

[GDC Download Tool](https://github.com/NCI-GDC/gdc-client/releases/latest)
