---
title: "Introduction to Accessing the NCI Genomic Data Commons"
author: "Sean Davis & Martin Morgan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
    df_print: paged
      
vignette: >
  %\VignetteIndexEntry{Introduction to Accessing the NCI Genomic Data Commons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, results='hide', echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
```

```{r libraries, message=FALSE}
library(GenomicDataCommons)
library(DT)
```

# What is the GDC?

From the [Genomic Data Commons (GDC) website](https://gdc.nci.nih.gov/about-gdc):

The National Cancer Institute's (NCI's) Genomic Data Commons (GDC) is
a data sharing platform that promotes precision medicine in
oncology. It is not just a database or a tool; it is an expandable
knowledge network supporting the import and standardization of genomic
and clinical data from cancer research programs.

The GDC contains NCI-generated data from some of the largest and most
comprehensive cancer genomic datasets, including The Cancer Genome
Atlas (TCGA) and Therapeutically Applicable Research to Generate
Effective Therapies (TARGET). For the first time, these datasets have
been harmonized using a common set of bioinformatics pipelines, so
that the data can be directly compared.

As a growing knowledge system for cancer, the GDC also enables
researchers to submit data, and harmonizes these data for import into
the GDC. As more researchers add clinical and genomic data to the GDC,
it will become an even more powerful tool for making discoveries about
the molecular basis of cancer that may lead to better care for
patients.

## GDC Data Model

The
[data model for the GDC is complex](https://gdc.cancer.gov/developers/gdc-data-model/gdc-data-model-components),
but it worth a quick overview. The data model is encoded as a
so-called property graph. Nodes represent entities such as Projects,
Cases, Diagnoses, Files (various kinds), and Annotations. The
relationships between these entities are maintained as edges.  Both
nodes and edges may have Properties that supply instance details.  The
GDC API exposes these nodes and edges in a somewhat simplified set
of
[RESTful](https://en.wikipedia.org/wiki/Representational_state_transfer) endpoints.

## Full Feature Set of the GDC API

HTTP interface that uses JSON as the primary data exchange format

* Programmatic access to functionality provided by GDC Data and
  Submission portals, via projects, cases, files, annotations, data,
  slicing, status, and submission endpoints
* Programmatic access to GDC Legacy Archive via legacy endpoint
* Token-based authentication for secure access to controlled data and
  to submission functionality
* RESTful search that supports simple and complex queries via filters,
  fields, and facets parameters, and project, files, cases, and
  annotations endpoints.
* Search results can be sorted using sort parameter, paginated using
  size and from parameters, and output in JSON, TSV, and XML using
  format and pretty parameters.
* _mapping endpoint enables user discovery of fields available for
  data search and retrieval operations
* Support for downloading of individual files and of archives
  containing multiple files
* [NOT Implemented Here] Generation of download and upload manifests
  for use with the GDC Data Transfer Tool
* BAM slicing functionality for downloading part(s) of a BAM file
  specified using chromosomal coordinates or HGNC gene names
* [NOT Implemented Here] Transactional submission system that links
  individual data elements according to a graph-based GDC Data Model
* Two data entity identifiers: UUIDs, which are consistent across GDC,
  and Submitter IDs, for compatibility with submitters’ tracking
  systems

Here, we will be describing how to access the main API endpoints:

- [Projects](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Projects/)
- [Files](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Cases_and_Files/)
- [Cases](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Cases_and_Files/)
- [Annotations](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Annotations/)
- [BAM Slicing](https://docs.gdc.cancer.gov/Data_Portal/Users_Guide/Cases_and_Files/#bam-slicing) - Specific to dealing with BAM files
- _mappings - For obtaining metadata about fields associated with the
  various endpoints


